---
title: "Practical Machine Learning - Prediction Assignment"
author: "Debadeep"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE)
```

## Background

Using devices like Jawbone Up, Nike FuelBand, and Fitbit, it's possible to collect
large amounts of data about personal activity. In this project, we use accelerometer
data from the belt, forearm, arm, and dumbell of 6 participants to predict the manner
in which they performed barbell lifts (the `classe` variable: A, B, C, D, E).

## Loading the Data

```{r load-packages}
library(caret)
library(randomForest)
library(dplyr)
```

```{r load-data}
train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

if (!file.exists("pml-training.csv")) download.file(train_url, "pml-training.csv")
if (!file.exists("pml-testing.csv"))  download.file(test_url, "pml-testing.csv")

training_raw <- read.csv("pml-training.csv", na.strings = c("NA", "", "#DIV/0!"))
testing_raw  <- read.csv("pml-testing.csv",  na.strings = c("NA", "", "#DIV/0!"))

dim(training_raw)
dim(testing_raw)
```

## Cleaning the Data

Three things need to go before modelling:

1. **Metadata columns** (row id, participant name, timestamps, window info) — these
   aren't real predictors and would leak information / add noise.
2. **Columns that are almost entirely NA** — many summary/statistic columns (avg,
   var, stddev, etc.) are blank except on "new window" rows. We drop any column
   with more than 95% missing values.
3. **Near-zero-variance columns** — columns that barely vary carry no signal.

```{r clean-data}
# Drop metadata columns (first 7 columns are identifiers/timestamps/window info)
training <- training_raw[, -(1:7)]
testing  <- testing_raw[, -(1:7)]

# Drop columns that are mostly NA (threshold: keep only if <5% missing)
na_ratio <- sapply(training, function(x) mean(is.na(x)))
keep_cols <- names(na_ratio[na_ratio < 0.05])

training <- training[, keep_cols]
# testing must keep the same predictor columns (minus classe, which testing lacks)
keep_cols_test <- setdiff(keep_cols, "classe")
testing <- testing[, keep_cols_test]

# Drop near-zero-variance columns (based on training set)
nzv <- nearZeroVar(training)
if (length(nzv) > 0) {
  training <- training[, -nzv]
  testing  <- testing[, intersect(names(testing), names(training))]
}

training$classe <- as.factor(training$classe)

dim(training)
dim(testing)
```

## Splitting into Training and Validation Sets

We hold out 30% of the labeled training data as a validation set. This lets us
estimate out-of-sample accuracy honestly, since the model never sees this slice
during training.

```{r partition}
set.seed(1234)
inTrain <- createDataPartition(training$classe, p = 0.7, list = FALSE)
train_set <- training[inTrain, ]
valid_set <- training[-inTrain, ]

dim(train_set)
dim(valid_set)
```

## Model Training with Cross-Validation

We use a **Random Forest**, which handles high-dimensional sensor data well and
is robust to noisy predictors. We use **5-fold cross-validation** inside the
training process itself (via `trainControl`) so the reported in-training accuracy
is already a cross-validated estimate, not just a single train/test split.

```{r train-model}
control <- trainControl(method = "cv", number = 5, allowParallel = TRUE)

set.seed(1234)
model_rf <- train(classe ~ ., data = train_set,
                   method = "rf",
                   trControl = control,
                   ntree = 150)

model_rf
```

## Evaluating on the Validation Set

This is where we get our real estimate of **out-of-sample error**, since
`valid_set` was never touched during training or cross-validation.

```{r validate}
pred_valid <- predict(model_rf, valid_set)
cm <- confusionMatrix(pred_valid, valid_set$classe)
cm

accuracy <- cm$overall["Accuracy"]
out_of_sample_error <- 1 - accuracy
```

**Out-of-sample accuracy:** `r round(accuracy * 100, 2)`%
**Estimated out-of-sample error:** `r round(out_of_sample_error * 100, 2)`%

## Why These Choices

- **Random Forest** was chosen over a single decision tree because it averages
  many trees trained on bootstrapped samples with random feature subsets,
  drastically reducing variance/overfitting compared to one tree — at some cost
  to interpretability, which is an acceptable trade-off here since the goal is
  pure predictive accuracy.
- **5-fold CV during training** protects against overfitting to any single
  train/validation split when tuning `mtry`.
- **Removing near-zero-variance and mostly-NA columns** avoids feeding the model
  noise and prevents errors from columns that don't exist meaningfully in the
  test set.
- The held-out **30% validation set** gives an honest, independent measure of
  out-of-sample error, separate from the cross-validation happening during
  training.

## Predicting the 20 Test Cases

```{r final-predictions}
final_predictions <- predict(model_rf, testing)
final_predictions
```

```{r write-answers}
# Helper to write each prediction to its own file (classic Coursera quiz format)
pml_write_files <- function(x) {
  n <- length(x)
  for (i in 1:n) {
    filename <- paste0("problem_id_", i, ".txt")
    write.table(x[i], file = filename, quote = FALSE,
                row.names = FALSE, col.names = FALSE)
  }
}
pml_write_files(final_predictions)
```

## Conclusion

The Random Forest model achieved an estimated out-of-sample accuracy of
`r round(accuracy * 100, 2)`% on the held-out validation set, cross-validated
with 5 folds during training. This model was then used to predict the 20
test cases for the quiz portion.
